Weight-adjusted discontinuous Galerkin methods: matrix-valued weights and elastic wave propagation in heterogeneous media

### 2016-12-21 docs/elasticity_notes/main.tex
local commits (not sure wtf this other stuff is)

\newcommand{\vect}[1]{\ensuremath\boldsymbol{#1}}
\newcommand{\tensor}[1]{\underline{\bm{#1}}}
\newcommand{\del}{\triangle}
\newcommand{\curl}{\grad \times}
\renewcommand{\div}{\grad \cdot}
\newcommand{\td}[2]{\frac{{\rm d}#1}{{\rm d}{\rm #2}}}
\newcommand{\pd}[2]{\frac{\partial#1}{\partial#2}}
\newcommand{\pdd}[2]{\frac{\partial^2#1}{\partial#2^2}}

\newcommand{\bs}[1]{\boldsymbol{#1}}

\newcommand{\equaldef}{\stackrel{\mathrm{def}}{=}}

\newcommand{\mi}{\mathrm{i}} %% roman "i"
\newcommand{\mb}[1]{\mathbf{#1}}
\newcommand{\mbb}[1]{\mathbb{#1}}
\newcommand{\mc}[1]{\mathcal{#1}}
\newcommand{\nor}[1]{\left\| #1 \right\|}
\newcommand{\LRp}[1]{\left( #1 \right)}
\newcommand{\LRs}[1]{\left[ #1 \right]}
\newcommand{\LRa}[1]{\left\langle #1 \right\rangle}
\newcommand{\LRb}[1]{\left| #1 \right|}
\newcommand{\LRc}[1]{\left\{ #1 \right\}}
\newcommand{\LRceil}[1]{\left\lceil #1 \right\rceil}
\newcommand{\LRl}[1]{\left. #1 \right|}

@@

\renewcommand{\Oh}{{\Omega_h}}
\renewcommand{\L}{L^2\LRp{\Omega}}
%\newcommand{\Lk}{\Lk}
%\newcommand{\Ldk}{L^2\LRp{\partial D^k}}
\newcommand{\Lk}{D^k}
\newcommand{\Ldk}{\partial D^k}
\newcommand{\Dhat}{\widehat{D}}
\newcommand{\Lhat}{L^2\LRp{\Dhat}}

@@

\date{}
\author{Jesse Chan}
\title{Weight-adjusted discontinuous Galerkin methods: matrix-valued weights and linear elastic wave propagation}

@@

\begin{abstract}
Weight-adjusted inner products \cite{chan2016weight1,chan2016weight2} replace weighted $L^2$ inner products with easily invertible weight-adjusted approximations.  When paired with discontinuous Galerkin (DG) methods, the result is a low storage, energy stable, and (for sufficiently regular weights) high order accurate method for wave propagation in the presence of general heterogeneous media and curvilinear meshes.  In this work, we extend weight-adjusted DG (WADG) methods to weighted inner products with matrix-valued weights, focusing on the linear elastic wave equation as an application.  We present a new DG formulation based on the discretization of the symmetric form of the linear elasticity wave equation, with upwind-like dissipation incorporated through simple penalty terms in the numerical fluxes.  This results in provably stable formulations for arbitrary anisotropic stiffness tensors and curvilinear meshes.  A convergence analysis is given, and numerical results confirm the stability and high order accuracy of WADG for several problems in elastic wave propagation.  
\end{abstract}

@@

There exist many high order energy stable discontinuous Galerkin formulations for the elastic wave equations in first order form.  Wilcox et al.\ introduced a discontinuous Galerkin spectral element method on hexahedral meshes

SEM \cite{komatitsch2010high}

DG-SEM \cite{wilcox2010high}

High order DG \cite{kaser2006arbitrary, dumbser2006arbitrary, de2007arbitrary, ye2016discontinuous}

Interior penalty formulations.  Analysis: Riviere et al.\ \cite{riviere2003discontinuous,riviere2007discontinuous} and Delcourte et al.\ \cite{delcourte2009high,delcourte2015analysis}

GPU-accelerated DG methods \cite{klockner2009nodal,godel2010gpu,godel2010scalability}.  Also elastic wave propagation and seismic wave applications \cite{modave2015nodal, modave2016gpu}

High order approximations of material data \cite{castro2010seismic, mercerat2015nodal}, but not low storage or not energy stable.

Extending mass lumping without dispersive effects \cite{christon1999influence,guermond2013correction}.  Reduces to exact mass matrix inversion for constant weights.

A simpler extension of WADG \cite{chan2016weight1,chan2016weight2} to elasticity and weight-adjusted inner products for vector-valued functions.

Outline: present elasticity and an energy stable DG method, motivate need for alternative to inverting large dense coupled mass matrix, introduce weight-adjusted inner products.

\section{Symmetric form of the elastic wave equation }

We begin with the linear elastic wave equation under small deformations in a domain $\Omega$.  These equations can be written a first order velocity-stress system for velocity $\bm{v}$ and symmetric stress tensor $\bm{S}$ 
\begin{align*}
\rho \pd{\bm{v}}{t} &= \Div{\bm{S}} + \bm{f}\\
\pd{\bm{S}}{t} &= \frac{1}{2} \bar{\bm{C}}\LRp{\Grad\bm{v} + \Grad\bm{v}^T},
\end{align*}
where $\bm{f}$ is the body force per unit volume, $\rho$ is density, and $\bar{\bm{C}}$ is the symmetric constitutive stiffness tensor relating stress and strain.  These equations can also be rewritten as a symmetric hyperbolic system of PDEs \cite{hughes1978classical}  
\begin{align}
\rho \pd{\bm{v}}{t} &= \sum_{i=1}^d \bm{A}_i^T \pd{\bm{\sigma}}{\bm{x}_i}\nonumber + \bm{f}\\
\bm{C}^{-1} \pd{\bm{\sigma}}{t} &= \sum_{i=1}^d \bm{A}_i \pd{\bm{v}}{\bm{x}_i}.
\label{eq:symelas}
\end{align}
where $\bm{C}$ is the symmetric matrix form of the constitutive tensor $\bar{\bm{C}}$ and $\sigma$ is a vector of length $N_d = \frac{d(d+1)}{2}$, the number of unique entries of the stress tensor $\bm{S}$.  We note that the matrices $\bm{A}_i$ are constant, while $\rho$ and $\bm{C}$ can vary spatially.  Furthermore, we will assume that $\rho$ and $\bm{C}$ are positive-definite and bounded such that 
\begin{align*}
\rho_{\min} &\leq \rho(\bm{x}) \leq \rho_{\max} \\
0 < c_{\min} \leq \bm{x}^T\bm{C}\bm{x} &\leq c_{\max} < \infty, \qquad \bm{x} \in \mathbb{R}^{N_d},
\end{align*}
which also implies that $\bm{C}^{-1}$ is positive-definite and bounded.

In two dimensions, $\bm{v} = (\bm{v}_1, \bm{v}_2)^T$ and $\bm{\sigma}= (\sigma_{xx},\sigma_{yy},\sigma_{xy})^T$ 
\[
\bm{S} = \LRp{\begin{array}{cc}
\sigma_{xx} & \sigma_{xy}\\
\sigma_{xy} & \sigma_{yy}
\end{array}}.
\]
The matrices $\bm{A}_i$ are given as
\[
\bm{A}_1 = \left(\begin{array}{cc}
1 & 0\\
0 & 0\\
0 & 1
\end{array}\right), \qquad 
\bm{A}_2 = \left(\begin{array}{cc}
0 & 0\\
0 & 1\\
1 & 0
\end{array}\right).
\]

In three dimensions, the velocity is $\bm{v} = (\bm{v}_1,\bm{v}_2,\bm{v}_3)^T$, while $\bm{\sigma}= (\sigma_{xx},\sigma_{yy},\sigma_{zz},\sigma_{yz},\sigma_{xy},\sigma_{xz})^T$ denotes the unique entries of the stress tensor $\bm{S}$
\[
\bm{S} = \LRp{\begin{array}{ccc}
\sigma_{xx} & \sigma_{xy} & \sigma_{xz}\\
\sigma_{xy} & \sigma_{yy}& \sigma_{yz}\\
\sigma_{xz} & \sigma_{yz}& \sigma_{zz}
\end{array}}.
\]
The matrices $\bm{A}_i$ are then
\[
\bm{A}_1 = \left(\begin{array}{ccc}
1 & 0 & 0\\
0 & 0& 0\\
0 & 0& 0\\
0 & 0& 0\\
0 & 0 & 1\\
0 & 1 & 0
\end{array}\right), \qquad 
\bm{A}_2 = \left(\begin{array}{ccc}
0 & 0 & 0\\
0 & 1 & 0\\
0 & 0 & 0\\
0 & 0 & 1\\
0 & 0 & 0\\
1 & 0 & 0
\end{array}\right), \qquad
\bm{A}_3 = \left(\begin{array}{ccc}
0 & 0 & 0\\
0 & 0 & 0\\
0 & 0 & 1\\
0 & 1 & 0\\
1 & 0 & 0\\
0 & 0 & 0
\end{array}\right)
\]

In general anisotropic media, $\bm{C}$ is symmetric and positive-definite.  For two-dimensional isotropic media, $\bm{C}$ is given as
\[
\bm{C}= \LRp{
\begin{array}{ccc}
2\mu+\lambda  &  \lambda  &     0\\
\lambda &       2\mu+\lambda &      0\\
          0 &      0           &     {\mu}
\end{array}}, 
\]
where $\lambda,\mu$ are Lame parameters.  For three-dimensional isotropic media, $\bm{C}$ is given instead by
\[
\bm{C}= \LRp{
\begin{array}{cccccc}
2\mu+\lambda  &  \lambda  &     \lambda  & & & \\
\lambda &   2\mu+\lambda & \lambda   & & & \\
\lambda & \lambda &   2\mu+\lambda &    & &\\
 &   &    & {\mu} & &\\
&  &   &    & {\mu} &\\
& &  &   &    & {\mu} \\
\end{array}}.  
\]
We will consider both spatially varying isotropic and anisotropic media in this work.

\section{An energy stable discontinuous Galerkin formulation for elastic wave propagation}

The advantage of the symmetric first order formulation of the elastic wave equations (\ref{eq:symelas}) is that it is straightforward to derive an energy stable discontinuous Galerkin formulation.  We assume that the domain $\Omega$ is Lipschitz and exactly triangulated by a mesh $\Omega_h$, which consists of elements $D^k$.  We further assume that each element $D^k$ is the image of a reference element $\widehat{D}$ under the local elemental mapping 
\[
\bm{x}^k = \bm{\Phi}^k \widehat{\bm{x}},
\]
where $\bm{x}^k = \LRc{x^k,y^k,z^k}$ denote physical coordinates on $D^k$ and $\widehat{\bm{x}} = \LRc{\widehat{x},\widehat{y},\widehat{z}}$ denote coordinates on the reference element.  We denote the determinant of the Jacobian of $\bm{\Phi}^k$ as $J$, and refer to it as the Jacobian for the remainder of this work.

We will approximate components over each element $D^k$ from an approximation space $V_h\LRp{D^k}$, which we define as the composition of the mapping $\bm{\Phi}^k$ and a reference approximation space $V_h\LRp{\widehat{D}}$
\[
V_h\LRp{D^k} = \bm{\Phi}^k \circ V_h\LRp{\widehat{D}}.
\]
For the remainder of this work, we will take $V_h\LRp{\widehat{D}} = P^N\LRp{\widehat{D}}$, where $P^N\LRp{\widehat{D}}$ is the polynomial space of total degree $N$ on the reference simplex.  In two dimensions, 
\[
P^N\LRp{\widehat{D}} = \LRc{ \widehat{x}^i \widehat{y}^j, \quad 0 \leq i + j \leq N},
\] 
and in three dimensions, 
\[
P^N\LRp{\widehat{D}} = \LRc{ \widehat{x}^i \widehat{y}^j \widehat{z}^k, \quad 0 \leq i + j +k \leq N}.  
\]

We denote the $L^2$ inner product and norm over $D^k$ by $\LRp{\cdot,\cdot}_{\Lk}$, such that
\[
\LRp{\bm{f},\bm{g}}_{\Lk} = \int_{D^k} \bm{f}\cdot\bm{g} \diff x = \int_{\widehat{D}} \bm{f}\cdot\bm{g} J, \qquad \nor{\bm{f}}_{\Lk}^2 = \LRp{\bm{f},\bm{f}}_{\Lk},
\]
where $\bm{f},\bm{g}$ are vector-valued functions.  Global $L^2$ inner products and norms are using local $L^2$ inner products and norms
\[
\LRp{\bm{f},\bm{g}}_{\Oh} = \sum_{D^k\in \Oh} \int_{D^k} \LRp{\bm{f},\bm{g}}_{\Lk}, \qquad \nor{\bm{f}}_{\Oh}^2 = \sum_{D^k\in \Oh} \nor{\bm{f}}^2_{\Lk}.
\]

@@

We can now specify a DG formulation for the linear elastic wave equation (\ref{eq:symelas}).  Symmetric hyperbolic systems readily admit a DG formulation based on penalty fluxes, as detailed in \cite{chan2016short}.  For the linear elastic wave equation in symmetric first order form, this formulation is given as
\begin{align}
\sum_{D^k\in \Oh} \LRp{\rho \pd{\bm{v}}{t},\bm{w}}_{\Lk} &= 
\sum_{D^k\in \Oh} \LRp{\LRp{ \sum_{i=1}^d \bm{A}_i^T\pd{\bm{\sigma}}{\bm{x}_i} + \bm{f},\bm{w}}_{\Lk}  + \LRa{\frac{1}{2}\bm{A}_n^T\jump{\bm{\sigma}} + \frac{\tau_v}{2}\bm{A}_n^T\bm{A}_n\jump{\bm{v}},\bm{w}}_{\Ldk}} \nonumber\\
\sum_{D^k\in \Oh}\LRp{\bm{C}^{-1} \pd{\bm{\sigma}}{t},\bm{q}}_{\Lk} &=
\sum_{D^k\in \Oh}\LRp{\LRp{\sum_{i=1}^d \bm{A}_i \pd{\bm{v}}{\bm{x}_i},\bm{q}}_{\Lk} + \LRa{\frac{1}{2}\bm{A}_n\jump{\bm{v}} + \frac{\tau_{\sigma}}{2}\bm{A}_n\bm{A}_n^T\jump{\bm{\sigma}},\bm{q}}_{\Ldk}},
\label{eq:dgform}
\end{align}
for all $\bm{w},\bm{q}\in V_h\LRp{\Omega_h}$.  Here, $\tau_v,\tau_\sigma \geq 0$ are penalty parameters and $\bm{A}_n$ is the normal matrix defined on a face $f$ as $\bm{A}_n = \sum_{i=1}^d \bm{n}_i \bm{A}_i$.  In two dimensions, $\bm{A}_n$ is 
\[
\bm{A}_n =  \LRp{\begin{array}{cc}
{n}_x & 0\\
 0 & {n}_y\\
{n}_y & {n}_x
\end{array}}.
\]
while in three dimensions, $\bm{A}_n$ is
\[
\bm{A}_n =  \LRp{\begin{array}{ccc}
{n}_x & 0 & 0\\
 0 & {n}_y & 0\\
0&  0 & {n}_z \\
0 &{n}_z & {n}_y\\
{n}_z & 0 & {n}_x\\
{n}_y & {n}_z & 0\\
\end{array}}.  
\]

@@

&\leq - \sum_{f \in\Gamma_h \setminus \partial \Omega} \int_{f} \frac{\tau_v}{2}\LRb{\bm{A}_n\jump{\bm{v}}}^2 + \frac{\tau_\sigma}{2}\LRb{\bm{A}_n^T\jump{\bm{\sigma}}}^2.  

@@

&\sum_{D^k\in \Oh} \frac{1}{2}\pd{}{t}\LRp{\LRp{\rho \bm{v},\bm{v}}_{\Lk} + \LRp{\bm{C}^{-1} \bm{\sigma},\bm{\sigma}}_{\Lk}} \leq - \sum_{f \in\Gamma_h \setminus \partial \Omega} \int_{f} \frac{\tau_v}{2}\LRb{\bm{A}_n\jump{\bm{v}}}^2 + \frac{\tau_\sigma}{2}\LRb{\bm{A}_n^T\jump{\bm{\sigma}}}^2\nonumber\\

@@

\section{Weight-adjusted inner products for matrix-valued weights}

>>>>

\newcommand{\vect}[1]{\ensuremath\boldsymbol{#1}}
\newcommand{\tensor}[1]{\underline{\bm{#1}}}
\newcommand{\del}{\triangle}
\newcommand{\curl}{\grad \times}
\renewcommand{\div}{\grad \cdot}

\newcommand{\bs}[1]{\boldsymbol{#1}}
\newcommand{\equaldef}{\stackrel{\mathrm{def}}{=}}

\newcommand{\td}[2]{\frac{{\rm d}#1}{{\rm d}{\rm #2}}}
\newcommand{\pd}[2]{\frac{\partial#1}{\partial#2}}
\newcommand{\pdd}[2]{\frac{\partial^2#1}{\partial#2^2}}
\newcommand{\mb}[1]{\mathbf{#1}}
\newcommand{\mbb}[1]{\mathbb{#1}}
\newcommand{\mc}[1]{\mathcal{#1}}
\newcommand{\nor}[1]{\left\| #1 \right\|}
\newcommand{\LRp}[1]{\left( #1 \right)}
\newcommand{\LRs}[1]{\left[ #1 \right]}
\newcommand{\LRa}[1]{\left\langle #1 \right\rangle}
\newcommand{\LRb}[1]{\left| #1 \right|}
\newcommand{\LRc}[1]{\left\{ #1 \right\}}
\newcommand{\LRceil}[1]{\left\lceil #1 \right\rceil}
\newcommand{\LRl}[1]{\left. #1 \right|}

@@

\newcommand{\Oh}{{\Omega_h}}
\renewcommand{\L}{L^2\LRp{\Omega}}
%\newcommand{\Lk}{L^2\LRp{D^k}}
%\newcommand{\Ldk}{L^2\LRp{\partial D^k}}
\newcommand{\Lk}{D^k}
\newcommand{\Ldk}{\partial D^k}
\newcommand{\Dhat}{\widehat{D}}
\newcommand{\Lhat}{L^2\LRp{\Dhat}}

@@

\date{}
\author{Jesse Chan}
\title{Weight-adjusted discontinuous Galerkin methods: matrix-valued weights and elastic wave propagation in heterogeneous media}

@@

\begin{abstract}
Weight-adjusted inner products \cite{chan2016weight1,chan2016weight2} replace weighted $L^2$ inner products with easily invertible weight-adjusted approximations.  When paired with discontinuous Galerkin (DG) methods, the result is a low storage, energy stable, and (for sufficiently regular weights) high order accurate method for wave propagation in the presence of general heterogeneous media and curvilinear meshes.  In this work, we extend weight-adjusted DG (WADG) methods to weighted inner products with matrix-valued weights, focusing on the linear elastic wave equation as an application.  We present a unified DG formulation for isotropic and anisotropic media based on the discretization of the symmetric form of the linear elasticity wave equation, with upwind-like dissipation incorporated through simple penalty terms in the numerical fluxes.  This formulation is provably stable for arbitrary stiffness tensors and curvilinear meshes.  A convergence analysis is given, and numerical results confirm the stability and high order accuracy of WADG for several problems in elastic wave propagation.  
\end{abstract}

@@

Efficient and accurate simulations of elastic wave propagation form the foundation for various seismic applications, including imaging and inversion techniques, as well as rupture and earthquake simulations.  The most common methods for the simulation of waves are finite difference methods \cite{virieux1986p}; however, finite element methods have garnered interest due to their low numerical dispersion and ability to accomodate geometrically flexible unstructured meshes.

Typical methods for time-domain wave propagation utilize explicit time stepping methods, since the hyperbolic partial differential equations (PDEs) governing wave propagation typically admit less restrictive stable timestep restrictions than parabolic equations.  However, finite element methods require the inversion of a global mass matrix when paired with explicit time integrators.  On hexahedral meshes, this can be avoided by using spectral element methods (SEM), which utilize mass lumping to produce a diagonal mass matrix \cite{komatitsch1998spectral}.  This global inversion can also be avoided through the use of discontinuous Galerkin (DG) methods, which result in a locally invertible block diagonal mass matrix.  Due to difficulties in extending mass-lumping techniques to tetrahedra, high order DG methods are often employed for seismic simulations which use simplicial meshes \cite{kaser2006arbitrary, dumbser2006arbitrary, de2007arbitrary,delcourte2009high,delcourte2015analysis,ye2016discontinuous}.  Such methods also lend themselves well to efficient implementations using Graphics Processing Units (GPUs) \cite{klockner2009nodal, modave2015nodal, modave2016gpu, chan2015gpu}.

Most high order DG methods on simplicial meshes assume that material coefficients are constant over each element, which can result in spurious reflections and the loss of high order accuracy \cite{castro2010seismic,mercerat2015nodal,bencomo2015discontinuous}.  This limitation can be overcome for DG methods by incorporating sub-element variation in media through weighted mass matrices, which results in an energy stable method \cite{mercerat2015nodal, bencomo2015discontinuous}.  For DG methods on hexahedral meshes, mass lumping reduces the construction of weighted mass matrices to entry-wise scalings of a diagonal mass matrix.  However, on tetrahedral meshes, this approach necessitates the pre-computation and storage of factorizations for each local mass matrix, which greatly increases both storage costs and data transferred at high orders of approximation.  These costs are especially problematic for accelerator architectures such as GPUs, which possess limited memory.  %Additionally, excessive storage costs can require reduced problem sizes and under-utilization of accelerator cards.

Storage costs associated with weighted mass matrices can be avoided by approximating weighted $L^2$ inner products using weight-adjusted inner products, which result in easily invertible approximations to weighted mass matrices \cite{chan2016weight1,chan2016weight2}.  For sufficiently regular weights, high order accuracy is also retained.  When paired with an energy stable DG formulation, weight-adjusted inner products result in weight-adjusted DG methods (WADG), which preserve energy stability and high order accuracy while retaining a low asymptotic storage cost.  Additionally, unlike mass-lumping techniques, WADG does not rely on the use of inexact quadrature rules, and reduces to the exact inversion of mass matrices for constant weights.

Weight-adjusted DG methods have been applied to both acoustic wave propagation in heterogeneous media and on curvilinear meshes \cite{chan2016weight1,chan2016weight2}.  Both cases involve scalar weighting functions.  In this work, we extend weight-adjusted inner products to matrix-valued weights, and demonstrate the usefulness of this generalization to elastic wave propagation in arbitrary media.  A new energy stable DG formulation is introduced based on the symmetric form of the elastic wave equations, with upwind-like numerical dissipation introduced through simple penalty fluxes \cite{warburton2013low,ye2016discontinuous}.

This work proceeds as follows.  Sections~\ref{sec:symelas} and \ref{sec:dgform} begin by presenting an energy stable DG formulation with simple penalty fluxes for the elastic wave equation in symmetric hyperbolic form, and discusses issues related to storage and inversion of local mass matrices for material coefficients with sub-element variations.  Section~\ref{sec:mwadg} extends weight-adjusted approximations to weighted $L^2$ inner products and mass matrices to the case of matrix-valued weights, and proves interpolation estimates which account for the regularity of the matrix weight.  These results are incorporated into a weight-adjusted DG method for the linear elastic wave equations in Section~\ref{sec:wadgelas}.  Finally, numerical results in Section~\ref{sec:numerical} demonstrate the accuracy of this method for several analytic solutions in linear elasticity.

\section{Symmetric form of the elastic wave equation }
\label{sec:symelas}
We begin with the linear elastic wave equation under small deformations in a domain $\Omega \in \mathbb{R}^d$.  These equations can be written a first order velocity-stress system for velocity $\bm{v}$ and symmetric stress tensor $\bm{S}$ 
\begin{align*}
\rho \pd{\bm{v}}{t} &= \Div{\bm{S}} + \bm{f}\\
\pd{\bm{S}}{t} &= \frac{1}{2} \tensor{\bm{C}}\LRp{\Grad\bm{v} + \Grad\bm{v}^T},
\end{align*}
where $\bm{f}$ is the body force per unit volume, $\rho$ is density, and $\tensor{\bm{C}}$ is the symmetric constitutive stiffness tensor relating stress and strain.  We rewrite these equations as a symmetric hyperbolic system of PDEs \cite{hughes1978classical}  
\begin{align}
\rho \pd{\bm{v}}{t} &= \sum_{i=1}^d \bm{A}_i^T \pd{\bm{\sigma}}{\bm{x}_i}\nonumber + \bm{f}\\
\bm{C}^{-1} \pd{\bm{\sigma}}{t} &= \sum_{i=1}^d \bm{A}_i \pd{\bm{v}}{\bm{x}_i}.
\label{eq:symelas}
\end{align}
where $\bm{C}$ is the symmetric matrix form of the constitutive tensor $\tensor{\bm{C}}$ and $\sigma$ is a vector of length $N_d = \frac{d(d+1)}{2}$, the number of unique entries of the stress tensor $\bm{S}$ in $d$ dimensions.  We note that the matrices $\bm{A}_i$ are constant, while $\rho$ and $\bm{C}$ can vary spatially.  Furthermore, we will assume that $\rho$ and $\bm{C}$ are positive-definite and bounded pointwise such that 
\begin{align*}
0 < \rho_{\min} &\leq \rho(\bm{x}) \leq \rho_{\max} < \infty, \qquad \forall \bm{x} \in \mathbb{R}^d,\\
0 < c_{\min} \leq \bm{u}^T\bm{C}(\bm{x})\bm{u} &\leq c_{\max} < \infty, \qquad \forall \bm{u} \in \mathbb{R}^{N_d}, \quad \forall \bm{x} \in \mathbb{R}^d,
\end{align*}
which also implies that $\bm{C}^{-1}$ is pointwise positive-definite and bounded.

In two dimensions, $\bm{v} = (\bm{v}_1, \bm{v}_2)^T$ and $\bm{\sigma}= (\sigma_{xx},\sigma_{yy},\sigma_{xy})^T$ 
\[
\bm{S} = \LRp{\begin{array}{cc}
\sigma_{xx} & \sigma_{xy}\\
\sigma_{xy} & \sigma_{yy}
\end{array}},
\]
while the matrices $\bm{A}_i$ are 
\[
\bm{A}_1 = \left(\begin{array}{cc}
1 & 0\\
0 & 0\\
0 & 1
\end{array}\right), \qquad 
\bm{A}_2 = \left(\begin{array}{cc}
0 & 0\\
0 & 1\\
1 & 0
\end{array}\right).
\]
In three dimensions, the velocity is $\bm{v} = (\bm{v}_1,\bm{v}_2,\bm{v}_3)^T$, while $\bm{\sigma}= (\sigma_{xx},\sigma_{yy},\sigma_{zz},\sigma_{yz},\sigma_{xy},\sigma_{xz})^T$ denotes the unique entries of the stress tensor $\bm{S}$
\[
\bm{S} = \LRp{\begin{array}{ccc}
\sigma_{xx} & \sigma_{xy} & \sigma_{xz}\\
\sigma_{xy} & \sigma_{yy}& \sigma_{yz}\\
\sigma_{xz} & \sigma_{yz}& \sigma_{zz}
\end{array}}.
\]
The matrices $\bm{A}_i$ are then
\[
\bm{A}_1 = \left(\begin{array}{ccc}
1 & 0 & 0\\
0 & 0& 0\\
0 & 0& 0\\
0 & 0& 0\\
0 & 0 & 1\\
0 & 1 & 0
\end{array}\right), \qquad 
\bm{A}_2 = \left(\begin{array}{ccc}
0 & 0 & 0\\
0 & 1 & 0\\
0 & 0 & 0\\
0 & 0 & 1\\
0 & 0 & 0\\
1 & 0 & 0
\end{array}\right), \qquad
\bm{A}_3 = \left(\begin{array}{ccc}
0 & 0 & 0\\
0 & 0 & 0\\
0 & 0 & 1\\
0 & 1 & 0\\
1 & 0 & 0\\
0 & 0 & 0
\end{array}\right)
\]

In general anisotropic media, $\bm{C}$ is symmetric and positive-definite.  For two-dimensional isotropic media, $\bm{C}$ is given as
\[
\bm{C}= \LRp{
\begin{array}{ccc}
2\mu+\lambda  &  \lambda  &     0\\
\lambda &       2\mu+\lambda &      0\\
          0 &      0           &     {\mu}
\end{array}}, 
\]
where $\lambda,\mu$ are Lame parameters.  For three-dimensional isotropic media, $\bm{C}$ is given instead by
\[
%\bm{C}= 
%\LRp{
%\begin{array}{cccccc}
%2\mu+\lambda  &  \lambda  &     \lambda  & & & \\
%\lambda &   2\mu+\lambda & \lambda   & & & \\
%\lambda & \lambda &   2\mu+\lambda &    & &\\
% &   &    & {\mu} & &\\
%&  &   &    & {\mu} &\\
%& &  &   &    & {\mu} \\
%\end{array}}.  
\LRp{\begin{array}{cccc}
2\mu+\lambda  &  \lambda  &     \lambda   & \\
\lambda &   2\mu+\lambda & \lambda   &  \\
\lambda & \lambda &   2\mu+\lambda &    \\
 &   &    & {\mu}\bm{I}^{3\times3} \\
\end{array}}.  
\]
We will consider both spatially varying isotropic and anisotropic media in this work.

\section{An energy stable discontinuous Galerkin formulation for elastic wave propagation}
\label{sec:dgform}
The advantage of the symmetric first order formulation of the elastic wave equations (\ref{eq:symelas}) is that it is straightforward to derive an energy stable discontinuous Galerkin formulation.  We assume that the domain $\Omega$ is Lipschitz and exactly triangulated by a mesh $\Omega_h$, which consists of elements $D^k$.  We further assume that each element $D^k$ is the image of a reference element $\widehat{D}$ under the local elemental mapping 
\[
\bm{x}^k = \bm{\Phi}^k \widehat{\bm{x}},
\]
where $\bm{x}^k = \LRc{x^k,y^k,z^k}$ denote physical coordinates on $D^k$ and $\widehat{\bm{x}} = \LRc{\widehat{x},\widehat{y},\widehat{z}}$ denote coordinates on the reference element.  We denote the determinant of the Jacobian of $\bm{\Phi}^k$ as $J$, and refer to it as the Jacobian for the remainder of this work.

We will approximate components over each element $D^k$ from an approximation space $V_h\LRp{D^k}$, which we define as the composition of the mapping $\bm{\Phi}^k$ and a reference approximation space $V_h\LRp{\widehat{D}}$
\[
V_h\LRp{D^k} = \bm{\Phi}^k \circ V_h\LRp{\widehat{D}}.
\]
For the remainder of this work, we will take $V_h\LRp{\widehat{D}} = P^N\LRp{\widehat{D}}$, where $P^N\LRp{\widehat{D}}$ is the polynomial space of total degree $N$ on the reference simplex.  In two dimensions, 
\[
P^N\LRp{\widehat{D}} = \LRc{ \widehat{x}^i \widehat{y}^j, \quad 0 \leq i + j \leq N},
\] 
and in three dimensions, 
\[
P^N\LRp{\widehat{D}} = \LRc{ \widehat{x}^i \widehat{y}^j \widehat{z}^k, \quad 0 \leq i + j +k \leq N}.  
\]
We denote the $L^2$ inner product and norm over $D^k$ by $\LRp{\cdot,\cdot}_{\Lk}$, such that
\[
\LRp{\bm{f},\bm{g}}_{\Lk} = \int_{D^k} \bm{f}\cdot\bm{g} \diff x = \int_{\widehat{D}} \bm{f}\cdot\bm{g} J, \qquad \nor{\bm{f}}_{\Lk}^2 = \LRp{\bm{f},\bm{f}}_{\Lk},
\]
where $\bm{f},\bm{g}$ are vector-valued functions.  Global $L^2$ inner products and norms are using local $L^2$ inner products and norms
\[
\LRp{\bm{f},\bm{g}}_{\Oh} = \sum_{D^k\in \Oh} \int_{D^k} \LRp{\bm{f},\bm{g}}_{\Lk}, \qquad \nor{\bm{f}}_{\Oh}^2 = \sum_{D^k\in \Oh} \nor{\bm{f}}^2_{\Lk}.
\]
We define also the $L^2$ inner product and norm over the boundary $\partial D^k$ of an element
\[
\LRa{u,v}_{\Ldk} = \int_{\partial D^k}uv\diff{x} = \sum_{f\in \partial D^k}\int_{\widehat{f}}uvJ^f\diff{\widehat{x}}, \qquad \nor{u}_{\Ldk}^2 = \LRa{u,u},
\]
where $J^f$ is the Jacobian of the mapping from a reference face (triangle) $\widehat{f}$ to a physical face $f$ of an element.

@@

We can now specify a unified DG formulation for the linear elastic wave equation (\ref{eq:symelas}) in either isotropic or anisotropic media.  Symmetric hyperbolic systems readily admit a DG formulation based on penalty fluxes \cite{chan2016short}.  For the linear elastic wave equation in symmetric first order form, this formulation is given as
\begin{align}
\sum_{D^k\in \Oh} \LRp{\rho \pd{\bm{v}}{t},\bm{w}}_{\Lk} &= 
\sum_{D^k\in \Oh} \LRp{\LRp{ \sum_{i=1}^d \bm{A}_i^T\pd{\bm{\sigma}}{\bm{x}_i} + \bm{f},\bm{w}}_{\Lk}  + \LRa{\frac{1}{2}\bm{A}_n^T\jump{\bm{\sigma}} + \frac{\tau_v}{2}\bm{A}_n^T\bm{A}_n\jump{\bm{v}},\bm{w}}_{\Ldk}} \nonumber\\
\sum_{D^k\in \Oh}\LRp{\bm{C}^{-1} \pd{\bm{\sigma}}{t},\bm{q}}_{\Lk} &=
\sum_{D^k\in \Oh}\LRp{\LRp{\sum_{i=1}^d \bm{A}_i \pd{\bm{v}}{\bm{x}_i},\bm{q}}_{\Lk} + \LRa{\frac{1}{2}\bm{A}_n\jump{\bm{v}} + \frac{\tau_{\sigma}}{2}\bm{A}_n\bm{A}_n^T\jump{\bm{\sigma}},\bm{q}}_{\Ldk}},
\label{eq:dgform}
\end{align}
for all $\bm{w},\bm{q}\in V_h\LRp{\Omega_h}$.  Here, $\tau_v,\tau_\sigma \geq 0$ are penalty parameters and $\bm{A}_n$ is the normal matrix defined on a face $f$ as $\bm{A}_n = \sum_{i=1}^d \bm{n}_i \bm{A}_i$.  In two dimensions, $\bm{A}_n$ is 
\[
\bm{A}_n =  \LRp{\begin{array}{cc}
{n}_x & 0\\
 0 & {n}_y\\
{n}_y & {n}_x
\end{array}}.
\]
while in three dimensions, $\bm{A}_n$ is
\[
\bm{A}_n =  \LRp{\begin{array}{ccc}
{n}_x & 0 & 0\\
 0 & {n}_y & 0\\
0&  0 & {n}_z \\
0 &{n}_z & {n}_y\\
{n}_z & 0 & {n}_x\\
{n}_y & {n}_z & 0\\
\end{array}}.  
\]

@@

&= - \sum_{f \in\Gamma_h \setminus \partial \Omega} \int_{f} \frac{\tau_v}{2}\LRb{\bm{A}_n\jump{\bm{v}}}^2 + \frac{\tau_\sigma}{2}\LRb{\bm{A}_n^T\jump{\bm{\sigma}}}^2.  

@@

&\sum_{D^k\in \Oh} \frac{1}{2}\pd{}{t}\LRp{\LRp{\rho \bm{v},\bm{v}}_{\Lk} + \LRp{\bm{C}^{-1} \bm{\sigma},\bm{\sigma}}_{\Lk}} = - \sum_{f \in\Gamma_h \setminus \partial \Omega} \int_{f} \frac{\tau_v}{2}\LRb{\bm{A}_n\jump{\bm{v}}}^2 + \frac{\tau_\sigma}{2}\LRb{\bm{A}_n^T\jump{\bm{\sigma}}}^2\nonumber\\

@@

\section{Weight-adjusted inner products for matrix-valued weights}
\label{sec:mwadg}
Weight-adjusted inner products are high order accurate approximations of weighted $L^2$ inner products.  These can be interpreted as generalizations of mass lumping techniques, reducing to mass lumping when integrals are evaluated with appropriate quadrature rules.  These weight-adjusted inner products result in weight-adjusted mass matrices, whose inverses approximate the inverse of a weighted $L^2$ mass matrix.
